% ........................................................................... %

\section*{Acknowledgments}

The story behind this PhD thesis started in 2004 while a Master by Research student at the Université Blaise Pascal in Clermont-Ferrand. As I was looking for an internship topic, Michel Schneider pointed me to Farouk Toumani, who would eventually become my PhD adviser. Thank you Michel for that, as Farouk proved to be a very good adviser. During those 4 years working on this topic, he always provided good guidance and support. It has been a great pleasure to work with him, both from a professional and a personal point of view. In turn, Farouk pointed me to Boualem Benatallah, who would become my PhD adviser at the University of New South Wales as I did my PhD \emph{under cotutelle agreements} between the two universities. I enjoyed working with Boualem very much, and his advices turned out to complement nicely those from Farouk.

I would like to thank the people with whom I shared the office room: Yoan Renaud that I first met when I entered the university, Alain Gély and Olivier Coupelon. We have had some great time together, and I am sure that there is still a lot more to come. There is a special mention for a group of 4 students of mine who later became my friends: Pierre Colomb, Olivier Montagner, Olivier Passalacqua and Frédéric Desforges. Of course I have enjoyed working with many other colleagues and friends in Clermont-Ferrand: Ramy Ragab, Marie Agier, Olivier Raynaud, Frédéric Flouvat, Hélène Jaudoin, Stéphanie Chollet, Kevin Hivernat, Yannick Loiseau, Raoul Médina, Béatrice Bourdieu and Françoise Toledo.

I would like to thank my coauthor Fabio Casati for the good times we have had working together, and for his never ending positive attitude. I met a lot of great people during my visits at UNSW. Warm thanks go to my friend Hamid Motahari and his wife Razieh for their kindness and for making my discovery of life in Australia so much easier. Among the people I met there, I would like to thank Halvard Skogsrud and Woralak Kongdenfha. Special thanks go to Mohand-Said Hacid that I first met while in Sydney. We spent some great time at UNSW, and he taught me a lot regarding running training!

Many thanks go to my reviewers and examiners as part of the French side of my PhD defense: Marie-Christine Fauvet, Schahram Dustdar, Claude Godart, Marlon Dumas and Michel Schneider. It has been a great honor to have you all as part of the examination committee!

My warmest thanks naturally go to my family, my parents Evelyne and Jean for having always stood behind me through the good and the bad times. I would probably not have gone so far in my studies without their encouragements and the education that they gave me. My final -- and biggest -- thanks go to Marie-Anne for her love and steady support during all of these years. Doing this PhD has required some sacrifices and I am deeply grateful to her not only because she accepted them, but because she fully and wholeheartedly endorsed them.

% ........................................................................... %