% ........................................................................... %

\paragraph{Abstract.}

Web services are increasingly gaining acceptance as a framework for facilitating
application-to-application interactions within and across enterprises. It is commonly accepted that a service description should include not only the interface, but also the \emph{business protocol} supported by the service. The present work focuses on the formalization of the important category of protocols that include time-related constraints (called \emph{timed protocols}), and the impact of time on compatibility and replaceability analysis.

We formalized the following timing constraints: \CInvoke constraints define time windows of availability while \MInvoke constraints define expirations deadlines. We extended techniques for compatibility and replaceability analysis between timed protocols by using a semantic-preserving mapping between timed protocols and timed automata, leading to the novel class of \emph{protocol timed automata} (PTA). Specifically, PTA exhibit silent transitions that cannot be removed in general, yet they are closed under complementation, making every type of compatibility or replaceability analysis decidable. Finally, we implemented our approach in the context of a larger project called ServiceMosaic, a model-driven framework for web service life-cycle management.

\paragraph{Keywords.}
Web services; timed business protocols; compatiblity and replaceability analysis; choreography; orchestration; business process; service-oriented architectures; timed automata.

% ........................................................................... %